---
title: "climate change and urban planning"
author: "Shuhan Song"
date: "12/3/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

#### Statistical Model for estimation of flood risk in NYC

**Assumptions**

- local sea level is affected only by tidal variations, storm surge, and changes to the long-term mean.

- use different fixed values of the tide level and storm surge

- all other sea level variations follow a normal (Gaussian distribution)

**Key variables** (meter)

current corrent normal sea level = 0

1. tidal variations

    - zero point is halfway between the high and low tide level
  
    - ranges from -1.25 to 1.25 m
  
    - use high tide level 1.25m and high tide level -1.25 m to model flood risks

2. storm surge

    - 1 m 
  
    - 2 m 
  
    - 3 m

3. changes to the long-term mean in 2050, which is caused by climate change (combining RCP 4.5 and RCP 8.5)

    - Based on the report by New York City Panel on Climate Change (NPCC) which predicted RCP 4.5 and RCP 8.5 scenarios in 2050s
  
    - low estimate (10th percentile): 7 inches
  
    - middle estimate (25-75th percentile): 11-24 inches
  
    - high estimate (90th percentile): 31 inches

4. background variations

    - normally distributed with mean = 0 and stdv = 3

**Criteria for Flood event**

Total sea level anomaly greater than 5m.

Total sea level anomaly = tide + storm surge + climate variability (background + climate change)


#### Model the model!
```{r}
library(tidyverse)
library(kableExtra)
library(here)
```


```{r}
# set up the variables
tide_level <- c(-1.25, 1.25)

storm_surge <- c(0, 1, 2, 3)

mean_2050 <- c(7, 11, 24, 31) * 0.0254 # conversion from inches to meters
```

#### Current condition:

```{r}
sea_level_cur <- 1
tide_level_cur <- 1
storm_surge_cur <- 1
true_mean_cur <- 1
n <- 5000
m <- 1

for (j in 1:length(storm_surge)) {
  for(t in 1:length(tide_level)){
    for (i in 1:n){
      sea_level_cur[m] <- storm_surge[j] + tide_level[t] + rnorm(1, mean = 0 , sd = 3)
      tide_level_cur[m] <- tide_level[t]
      storm_surge_cur[m] <- storm_surge[j]
      true_mean_cur[m] <- storm_surge[j] + tide_level[t]
      m = m+1
      }
    }
  }

flood_cur <- data_frame(tide_level_cur, storm_surge_cur, sea_level_cur, true_mean_cur)
```

```{r}
# test the model
test_cur <- flood_cur %>% 
  group_by(tide_level_cur, storm_surge_cur) %>% 
  summarize(mean = mean(sea_level_cur), 
            true_mean = mean(true_mean_cur)
            )

kable(test_cur,
      col.names = c("tide level", "storm surge", "mean", "true mean")) %>% 
  kable_styling(bootstrap_options = "striped")
```

```{r}
# flood risk
cur_flood_risk <- flood_cur %>% 
  filter(sea_level_cur > 5) %>% 
  group_by(tide_level_cur, storm_surge_cur) %>% 
  summarize(flood_risk_cur = n()/n,
            true_mean = mean(true_mean_cur))

kable(cur_flood_risk,
      col.names = c("tide level", "storm surge", "flood risk", "true mean")) %>% 
  kable_styling(bootstrap_options = "striped")
```

- The risk of flooding under normal conditions:
    - low tide: `r cur_flood_risk[1, 3]*100`% of flood risk
    - high tide: `r cur_flood_risk[5, 3]*100`% of flood risk

- The risk of flooding with storm surge:
    - 1m: 
        - low tide: `r cur_flood_risk[2, 3]*100`% of flood risk
        - high tide: `r cur_flood_risk[6, 3]*100`% of flood risk
        
    - 2m: 
        - low tide: `r cur_flood_risk[3, 3]*100`% of flood risk
        - high tide: `r cur_flood_risk[7, 3]*100`% of flood risk
    
    - 3m: 
        - low tide: `r cur_flood_risk[4, 3]*100`% of flood risk
        - high tide: `r cur_flood_risk[8, 3]*100`% of flood risk

**visualize the flood risk**

```{r}
risk_high_cur <- cur_flood_risk %>% 
  filter(tide_level_cur == 1.25)
risk_low_cur <- cur_flood_risk %>% 
  filter(tide_level_cur == -1.25)
  
risk_range_cur <- data_frame("storm_surge_cur" = risk_high_cur$storm_surge_cur,
                             "high_tide_risk" = risk_high_cur$flood_risk_cur,
                             "low_tide_risk" = risk_low_cur$flood_risk_cur)

ggplot(data = risk_range_cur,
       aes(x = storm_surge_cur)) +
  geom_point(aes(y = high_tide_risk),
             color = "darkblue") +
  geom_point(aes(y = low_tide_risk),
             color = "lightblue") +
  geom_line(aes(y = high_tide_risk),
             color = "darkblue") +
  geom_line(aes(y = low_tide_risk),
             color = "lightblue") +
  geom_ribbon(aes(ymin = low_tide_risk,
                  ymax = high_tide_risk),
              fill = "blue",
              alpha = 0.05) +
  labs(x = "Storm Surge (meters)",
       y = "Flood Risk",
       title = "Flood risk without climate change in NYC") +
  geom_text(x = 2.5, y = 0.08,
            label = "Low tide",
            color = "lightblue") +
  geom_text(x = 1, y = 0.22,
            label = "High tide",
            color = "darkblue") +
  geom_text(x = 2.5, y = 0.2,
            label = "Flood risk region",
            color = "blue") +
  theme_minimal()

ggsave(here::here("Current flood risk.jpg"), width = 6, height = 4)
```
        
#### Climate change in 2050

```{r}
sea_level_2050 <- 1
tide_level_2050 <- 1
storm_surge_2050 <- 1
mean_sea_level_2050 <- 1
true_mean_2050 <- 1
n <- 5000
m <- 1
for (k in 1:length(mean_2050)) {
  for (j in 1:length(storm_surge)) {
    for(t in 1:length(tide_level)){
      for (i in 1:n){
        sea_level_2050[m] <- storm_surge[j] + tide_level[t] + rnorm(1, mean = mean_2050[k] , sd = 3)
        tide_level_2050[m] <- tide_level[t]
        storm_surge_2050[m] <- storm_surge[j]
        mean_sea_level_2050[m] <- mean_2050[k]
        true_mean_2050[m] <- storm_surge[j] + tide_level[t] + mean_2050[k]
        m = m+1
        }
      }
    }
  }

flood_2050 <- data_frame(mean_sea_level_2050, tide_level_2050, storm_surge_2050, sea_level_2050, true_mean_2050)
```

```{r}
# test the model
test_2050 <- flood_2050 %>% 
  group_by(mean_sea_level_2050, tide_level_2050, storm_surge_2050) %>% 
  summarize(mean = mean(sea_level_2050), 
            true_mean = mean(true_mean_2050)
            )
kable(test_2050,
      col.names = c("mean sea leve", "tide level", "storm surge", "mean", "true mean")) %>% 
  kable_styling(bootstrap_options = "striped")
```


```{r}
# flood risk
flood_risk_2050 <- flood_2050 %>% 
  filter(sea_level_2050 > 5) %>% 
  group_by(mean_sea_level_2050, tide_level_2050, storm_surge_2050) %>% 
  summarize(flood_risk = n()/n,
            true_mean = mean(true_mean_2050))

kable(flood_risk_2050,
      col.names = c("mean sea leve", "tide level", "storm surge", "flood risk", "true mean")) %>% 
  kable_styling(bootstrap_options = "striped")
```
